
---
title: "WGCNA_final_code"
output: html_notebook
---
 
 
 
# Install BiocManager if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
 
# Install DESeq2
BiocManager::install("DESeq2")
```{r}
library(dplyr)
library(DESeq2)
library(WGCNA)
```
#1. Load and preprocess data

#using VST : is preferred for WGCNA because it provides a more statistically sound and stable input, improving the reliability of gene co-expression networks. 
```{r}

#==============================================================================
# Step 1: load the files and merge together raw read counts files
#===============================================================================
# Folder with your RPKM files, e.g. sample1.txt, sample2.txt ...
setwd("~/Downloads/WGCNA_Amanpreet/files")

files <- list.files(pattern = "\\.read_count$")
files

expr_list <- lapply(files, function(f) {
  df <- read.table(f, header = TRUE, sep = "\t")
  # Rename count column to sample name (file basename)
  sample_name <- tools::file_path_sans_ext(basename(f))
  colnames(df) <- c("GeneID", sample_name)
  return(df)
})

# Merge all data frames by GeneID
expr_merged <- Reduce(function(x, y) full_join(x, y, by = "GeneID"), expr_list)

# Replace NAs with 0 (genes not detected in some samples)
expr_merged[is.na(expr_merged)] <- 0

# Set rownames as GeneID and remove GeneID column
rownames(expr_merged) <- expr_merged$GeneID
expr_merged$GeneID <- NULL

# Check matrix dimensions (genes x samples)
dim(expr_merged)

metadata <- read.csv("/Users/azk0151/Downloads/WGCNA_genes/metadata.csv", header = TRUE)
head(metadata)

#Make sure sample names in expr_merged columns and metadata$sampleID match:
all(colnames(expr_merged) %in% metadata$sampleID)  # should be TRUE

#Reorder metadata to match counts columns:
metadata <- metadata[match(colnames(expr_merged), metadata$sampleID), ]

#write.table(expr_merged, "~/Downloads/WGCNA_Amanpreet/expr_merged_raw.txt", sep = "\t", quote = FALSE)

expr_merged <- read.table("/Users/azk0151/Downloads/WGCNA_genes/RAW_files-selected/expr_merged_raw.txt", header = TRUE, row.names = 1, sep = "\t")

#==============================================================================
# Step 2: normalizing
#===============================================================================


# Create DESeq2 dataset
#dds <- DESeqDataSetFromMatrix(countData = expr_merged, colData = metadata, design = ~ 1) # 1 means across all samples irespective of treatment
dds <- DESeqDataSetFromMatrix(countData = expr_merged,
                              colData   = metadata,   # metadata must include "condition"
                              design    = ~ Condition)
# keep genes that have at least 10 counts in at least 3 samples
keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]

#1 means Intercept only, no groups modeled,	Just normalization for VST, not differential expression

# Modules show natural co-expression patterns across all samples, reflecting true biological relationships, unbiased by known groups. Ideal for discovering unknown relationships.
vsd <- vst(dds, blind = TRUE)

#Usually, set blind = FALSE so the transformation respects the experimental design and variability between samples properly. This avoids removing or flattening true biological differences that you want to capture in the network.

# Extract normalized expression matrix
vsd_mat <- assay(vsd)
colnames(vsd_mat) <- sub("\\.read_count$", "", colnames(vsd_mat))

# Transpose for WGCNA (samples x genes)
datExpr <- t(vsd_mat)

# Save normalized data (optional)
#write.table(datExpr, "/Users/austudent/Downloads/WGCNA_Amanpreet/WGCNA_input_VST_normalized_con.txt", sep = "\t", quote = FALSE)

```

#1. Load and preprocess data
 
```{r}
#==============================================================================
# Step: WGCNA
#===============================================================================



library(WGCNA)
options(stringsAsFactors = FALSE)
enableWGCNAThreads()
# #/Users/azk0151/Downloads/WGCNA_genes/RAW_files-selected/WGCNA_input_VST_normalized.tx

# Read the gene counts table

#datExpr <- read.table("/Users/azk0151/Downloads/WGCNA_genes/RAW_files-selected/WGCNA_input_VST_normalized.txt", header = TRUE, row.names = 1, sep = "\t")

```

#2.  Sample clustering


```{r}
myOrder <- c("AAL6572HR1", "AAL6572HR2", "AAL6572HR3", "AAL6524HR1", "AAL6524HR2", "AAL6524HR3", "EAL6524HR1", "EAL6524HR2", "EAL6524HR3", "EAL658HR1", "EAL658HR2", "EAL658HR3"
             )

# Reorder expression & metadata
datExpr_manual <- datExpr[myOrder, ]
rownames(metadata) <- metadata$Accession_code
metadata_manual <- metadata[myOrder, ]

# Rebuild the tree using this fixed order
sampleTree <- hclust(dist(datExpr_manual), method = "single")

# Extract metadata columns
metadata2 <- metadata_manual[, c(4, 5)]

# **KEY FIX 1**: Don't reorder by sampleTree - use your manual order
# The whole point of myOrder was to control the order, so keep it!
datTraitsOrdered <- metadata2  # Already in your desired order

# Define color palettes
traitColorList <- list(
  Condition = c(
    "Epiphytic" = "#1f77b4",
    "Apoplastic" = "#d62728"
  ),
  Time = c(
    "8" = "#2ca02c",
    "24" = "#ff7f0e",
    "72" = "#9467bd"
  )
)

# **KEY FIX 2**: Convert factors to character for proper matching
traitColors <- data.frame(
  Condition = traitColorList$Condition[as.character(datTraitsOrdered$Condition)],
  Time = traitColorList$Time[as.character(datTraitsOrdered$Time)],
  row.names = rownames(datTraitsOrdered)
)

pdf(file = "/Users/azk0151/Downloads/sampleClusteringtree.pdf", width = 7, height = 5)

par(mar = c(10, 4, 4, 2))

plotDendroAndColors(
  sampleTree,
  colors = as.matrix(traitColors),
  groupLabels = colnames(traitColors),
  main = " ",
  dendroLabels = rownames(datExpr_manual),
  hang = 0.03,
  addGuide = FALSE,
  cex.dendroLabels = 0.8,
  cex.colorLabels = 0.8
)

# Use absolute positioning
par(xpd = TRUE)
usr <- par("usr")  # Get plot coordinates
legend(x = mean(usr[1:2]),  # Center horizontally
       y = usr[3] - 0.2 * diff(usr[3:4]),  # Position below plot
       legend = c("Epiphytic", "Apoplastic", "8hr", "24hr", "72hr"),
       fill = c("#1f77b4", "#d62728", "#2ca02c", "#ff7f0e", "#9467bd"),
       border = "black",
       bty = "n",
       horiz = TRUE,
       xjust = 0.5,  # Center the legend
       cex = 0.9)
par(xpd = FALSE)

dev.off()
```



```{r}
sampleTree = hclust(dist(datExpr), method = "single");
# plot sample tree
#pdf(file = "/Users/azk0151/Downloads/1-n-sampleClustering.pdf", width = 12, height = 6);
par(cex = 1.3);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="",
     cex.lab = 1.5,cex.axis = 1.5, cex.main = 2)
#dev.off()

```

#3. Soft-thresholding power
```{r}
#===============================================================================
#
#  Choose soft threshold parameter
#
#===============================================================================
 
# Choose a set of soft threshold parameters
powers = c(c(1:20), seq(from = 22, to=30, by=2))

sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Scale-free topology fit index as a function of the soft-thresholding power
#pdf(file = "2-n-sft.pdf", width = 9, height = 5);
par(mfrow = c(1,2));
cex1 = 0.8;
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
#abline(h = 0.80, col = "yellow", lty = 2)
#abline(h = 0.82, col = "orange", lty = 2)
#abline(h = 0.88, col = "pink", lty = 2)
abline(h = 0.80, col = "red", lty = 2)

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
#abline(h = 30, col = "yellow", lty = 2)
#abline(h = 100, col = "red", lty = 2)
#dev.off()

```



#step 4. Build adjacency and TOM
```{r}
#===============================================================================
#
#  Turn data expression into topological overlap matrix
#
#===============================================================================
 
# Turn data expression into topological overlap matrix

#7	0.826	-0.643	0.919	239.0	201.00	698

softPower <- 7
temp_cor <- cor
cor <- WGCNA::cor  # use WGCNA optimized correlation

# Adjacency matrix
adjacency <- adjacency(datExpr, power = softPower, type = "signed")

# signed:Keeps directionality.: Gives modules that are biologically more interpretable — all genes tend to “go up and down together.” 
#unsigned: mixed up and down gened within modules: Treats positive and negative correlations as equal.
# TOM (Topological Overlap Matrix)
TOM <- TOMsimilarity(adjacency, TOMType = "signed")
dissTOM <- 1 - TOM

cor <- temp_cor
```

#5. Hierarchical clustering of genes
```{r}
#===============================================================================
#
#  Construct modules (proceed with the genetree from option 2b)
#
#===============================================================================
# Plot gene tree
geneTree = hclust(as.dist(dissTOM), method = "average");
#pdf(file = "3-gene_cluster.pdf", width = 12, height = 9);
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04);
#dev.off()

```


#6.  Module detection (Dynamic Tree Cut)
```{r}
# Module identification using dynamic tree cut
# We like large modules, so we set the minimum module size relatively high:
# minModuleSize = 30;
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,deepSplit = 2,
                            pamRespectsDendro = FALSE,minClusterSize = 30);
table(dynamicMods)
length(table(dynamicMods))
# Convert numeric labels into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
#pdf(file = "4-module_tree.pdf", width = 8, height = 6);
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",dendroLabels = FALSE,
                    hang = 0.03,addGuide = TRUE, guideHang = 0.05,main = "Gene dendrogram and module colors")
#dev.off()
```

# 7. Calculate eigengenes and merge similar modules
```{r}
#===============================================================================
#
#  Merge modules
#
#===============================================================================
# Calculate eigengenes
MEList = moduleEigengenes(datExpr_manual, colors = dynamicColors)
MEs = MEList$eigengenes
MEs <- orderMEs(MEs)
head(MEs)

```

```{r}
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
sizeGrWindow(7, 6)
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
# Merge close modules
MEDissThres=0.25 # we can go as low  #This height corresponds to a correlation of over 75%. Any branches below this line are more than 75% related, and you will thus be merging them!
abline(h=MEDissThres, col = "red")
merge = mergeCloseModules(datExpr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
mergedColors = merge$colors 
mergedMEs = merge$newMEs 
# Plot merged module tree
#pdf(file = "5-merged_Module_Tree.pdf", width = 12, height = 9) 
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"), dendroLabels = FALSE,
                    hang = 0.03, addGuide = TRUE, guideHang = 0.05) 
#dev.off()
```

#8. Final module assignments

```{r}
moduleColors <- mergedColors
MEs <- mergedMEs
nSamples <- nrow(datExpr_manual)
modNames <- substring(names(MEs), 3)

#kME <- as.data.frame(cor(datExpr, MEs, use="p"))
geneModuleMembership <- as.data.frame(cor(datExpr_manual, MEs, use = "p"))
MMPvalue <- as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples))
names(geneModuleMembership) <- paste0("MM", modNames)
names(MMPvalue) <- paste0("p.MM", modNames)



geneInfo_condition <- data.frame(
  GeneID = colnames(datExpr),
  Module = moduleColors
) %>%
  cbind(geneModuleMembership) %>%
  cbind(MMPvalue)

```

```{r}
moduleColors <- mergedColors
MEs <- mergedMEs
module_order = names(MEs) %>% gsub("ME","", .)
# Add treatment names
MEs$treatment = row.names(MEs)

# tidy & plot data
library(tidyr)
library(ggplot2)


mME = MEs %>%
  pivot_longer(-treatment) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )


pdf(file = "/Users/azk0151/Downloads/module_trait.pdf", width = 9, height = 6)

mME %>% 
  ggplot(., aes(x = factor(treatment, 
                           levels = c("AAL6524HR1", "AAL6524HR2", "AAL6524HR3", 
                                     "AAL6572HR1", "AAL6572HR2", "AAL6572HR3", 
                                     "EAL6524HR1", "EAL6524HR2", "EAL6524HR3",
                                     "EAL658HR1", "EAL658HR2", "EAL658HR3")), 
                y = name, 
                fill = value)) +
  geom_tile(color = "white", linewidth = 0.5) +  # Add grid lines
  theme_bw() +
scale_fill_gradientn(
  colors = c("#0571B0", "white", "#CA0020"),
  values = scales::rescale(c(-1, 0, 0.8)),
  limit = c(-1, 0.8),
  breaks = seq(-1, 0.8, 0.2)) +
  labs(x = "Treatments", y = "Modules", fill = "Correlation", 
       title = " ") +  # Fixed labs() syntax
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

dev.off()
```



# for combined plot of dedragram and module-trait relatiosnhip plot
```{r}

library(ggplot2)
library(cowplot)
library(gridGraphics)

# ... (all your data preparation code remains the same) ...

myOrder <- c("AAL6572HR1", "AAL6572HR2", "AAL6572HR3", "AAL6524HR1", "AAL6524HR2", "AAL6524HR3", 
             "EAL6524HR1", "EAL6524HR2", "EAL6524HR3", "EAL658HR1", "EAL658HR2", "EAL658HR3")

datExpr_manual <- datExpr[myOrder, ]
rownames(metadata) <- metadata$Accession_code
metadata_manual <- metadata[myOrder, ]

sampleTree <- hclust(dist(datExpr_manual), method = "single")
metadata2 <- metadata_manual[, c(4, 5)]
datTraitsOrdered <- metadata2

traitColorList <- list(
  Condition = c("Epiphytic" = "#1f77b4", "Apoplastic" = "#d62728"),
  Time = c("8" = "#2ca02c", "24" = "#ff7f0e", "72" = "#9467bd")
)

traitColors <- data.frame(
  Condition = traitColorList$Condition[as.character(datTraitsOrdered$Condition)],
  Time = traitColorList$Time[as.character(datTraitsOrdered$Time)],
  row.names = rownames(datTraitsOrdered)
)

# Create Plot B (ggplot)
plot_B <- mME %>% 
  ggplot(., aes(x = factor(treatment, 
                           levels = c("AAL6524HR1", "AAL6524HR2", "AAL6524HR3", 
                                     "AAL6572HR1", "AAL6572HR2", "AAL6572HR3", 
                                      "EAL6524HR2", "EAL6524HR3", "EAL6524HR1",
                                     "EAL658HR1", "EAL658HR2", "EAL658HR3")), 
                y = name, 
                fill = value)) +
  geom_tile(color = "white", linewidth = 0.5) +
  theme_bw() +
  scale_fill_gradientn(
    colors = c("#0571B0", "white", "#CA0020"),
    values = scales::rescale(c(-1, 0, 0.8)),
    limit = c(-1, 0.8),
    breaks = seq(-1, 0.8, 0.2)) +
  labs(x = "Treatments", y = "Modules", fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 10))

# Function to create Plot A
plot_A_func <- function() {
  par(mar = c(10, 4, 4, 2))
  
  plotDendroAndColors(
    sampleTree,
    colors = as.matrix(traitColors),
    groupLabels = colnames(traitColors),
    main = " ",
    dendroLabels = rownames(datExpr_manual),
    hang = 0.03,
    addGuide = FALSE,
    cex.dendroLabels = 0.8,
    cex.colorLabels = 0.8
  )
  
  par(xpd = TRUE)
  usr <- par("usr")
  legend(x = mean(usr[1:2]),
         y = usr[3] - 0.2 * diff(usr[3:4]),
         legend = c("Epiphytic", "Apoplastic", "8hr", "24hr", "72hr"),
         fill = c("#1f77b4", "#d62728", "#2ca02c", "#ff7f0e", "#9467bd"),
         border = "white",
         bty = "n",
         horiz = TRUE,
         xjust = 0.5,
         cex = 0.9)
  par(xpd = FALSE)
}

# METHOD 1: Using pdf() directly with high resolution
pdf(file = "/Users/azk0151/Downloads/combined_plots.pdf", 
    width = 12, 
    height = 14)

# Combine using cowplot
combined_plot <- plot_grid(
  as_grob(plot_A_func),
  plot_B,
  ncol = 1,
  labels = c("A", "B"),
  label_size = 16,
  label_fontface = "bold",
  rel_heights = c(1, 1)
)

print(combined_plot)

dev.off()
```

